Graphs, partitioning and you

### 2016-01-31 evolutionaryComputation/2-practical/report.tex
added a bunch of text!

\newcommand{\brcell}[2][l]{%
	\begin{tabular}[#1]{@{}l@{}}#2\end{tabular}}
\begin{document}
\begin{empfile}
\begin{empcmds}
input metauml;
\end{empcmds}
\author{Jappie Klooster}
\title{Graph bipartitioning and you}
\maketitle

@@

\section{Assignment}
Graph bi-partitioning is dividing a graph into 2, with as little edges between
both graphs as possible.

@@

@@

I got much better results with a system where the local search got several chances.
If it failed it would fall back, and then try again.
With chances around 30 I managed to get up and around 1900.
When I reset the chance counter on a successful find
I managed to get consistently 2200, one result was as high as 2310.

I don't think there is a way to do it any better without drastically making
the algorithm more complex.
This current search for improvement was just based on the fact that ILS was
under performing to MLS, which shouldn't be the case according to my memory
from class. Now its over-performing quite drastically.

\subsection{GLS}
I implemented GLS with the evolution class. The only thing I had to do really
was write the crossover method, the rest was already functional.

@@

Here I also started to think about my own topic to investigate, at first
I just wanted to do varying population sizes and varying selection methods.
But I thinks this takes a little to much time and it doesn't quite fit
on the tTest table so I decided to drop the varying pop-sizes.

\subsection{Counting double}
I had an issue in which my classmates would get much better results than me.
After discussing it for a while it turned out that I was counting double. However
This didn't make sense because I just used Scala combinators to count the
elements. Basically this means that there would have to be a bug in the std
Scala library which would seem highly unlikely. I knew I was counting double
because all my results where even numbers. After thinking about it for a while
I decided to investigate the graph file, it turns out that every connection
is made double, so the double counting was in fact correct.

\subsubsection{Graph pruning}
I decided to use a prune graph to speed up the algorithm. In the beginning I
create 2 graphs. One normal and one pruned, The pruning works by selecting
comparing all connections against the id, if lower we ignore the connection.
To make sure we only do this on doubly linked graphs 2 graphs are created.
If the prune graph has exactly half the score of a random solution we assume
its a doubly linked graph so we can prune it.
The pruning effectively halves the amount of connections that need to be
checked by each iteration (this happens in the evaluation). The
Fiduccia Mattheyses has even more gain because it has a lot of book keeping to
do with the connections.

\subsection{Fiduccia Mattheyses}
I was somewhat sceptical of my local search implementation. I had quite 
different results from my classmates so to compare I decided to implement
the Fiduccia Mattheyses local search to. It was quite a lot of work but it
truly is significantly faster.

Finding the paper was somewhat difficult as the original was behind a paywall.
But someone put a copy on GitHub so I just used that one. The algorithm 
involves a lot of cleaver book keeping. First of all the concept of gain is
introduced, this indicates how much gain there will be from moving a vertex
to the other partition. Then the vertexes are put into an array with as index
their gain. If there are more vertices at a particular index then they're
put into an linked list. To support this I introduced a container class
Cell that holds the vertex and also has an optional previous and next
element.
These cells are also put in another array based on their vertex ID. This
array is used for constant speed access when the connections of a vertex their
gain needs to be updated after moving it to the other partition.
So you can just untie the element of the linked list (or if it has no
previous in my case its in the front of the bucket and I remove it from
there and reinsert the next element).

@@

As can be seen in the results section the Fiduccia Mattheyses algorithm is
much faster than first improvement vertex swap. The main reason for this is
that a single pass will climb very steeply in valuation. It also appears
to produce better results, I think this is because highest gain is considered
first.

My other research topic is selection pressure, turns out that tournament is
the best for GLS in case of Fiduccia Mattheyses, but it also takes the mot
amount of time.

\subsection{Statistics}
The error margins in the graph are the maximum range of the results so these
are round numbers when subtracted from the low or added to the high. The low
and high values are calculated with something called a confidence interval.
It aligns pretty well with the tTest results so I guess its OK.

@@

\end{tikzpicture}
\\
\input{vertResults.textable}
\end{empfile}
\end{document}

>>>>

\newcommand{\brcell}[2][l]{%
	\begin{tabular}[#1]{@{}l@{}}#2\end{tabular}}

% "define" Scala
\lstdefinelanguage{scala}{
  morekeywords={abstract,case,catch,class,def,%
    do,else,extends,false,final,finally,%
    for,if,implicit,import,match,mixin,%
    new,null,object,override,package,%
    private,protected,requires,return,sealed,%
    super,this,throw,trait,true,try,%
    type,val,var,while,with,yield},
  otherkeywords={=>,<-,<\%,<:,>:,\#,@},
  sensitive=true,
  morecomment=[l]{//},
  morecomment=[n]{/*}{*/},
  morestring=[b]",
  morestring=[b]',
  morestring=[b]"""
}
\begin{document}
\begin{empfile}
\begin{empcmds}
input metauml;
\end{empcmds}
\author{Jappie Klooster}
\title{Graphs, partitioning and you}
\maketitle

@@

\section{Assignment}
Graph bi-partitioning is dividing a graph into 2 partitions, 
with as little crossings between both partitionings as possible.

So we try to reduce the amount of crossing by modifiying which vertex resides
in what partition. The graph itself is however not modified. The amount of
vertexes that reside in on partition should also be even to the other partition.

@@

\newpage
\subsubsection{Reading the graph}
The first thing I did was reading the graph into a memory structure. That 
looks like this:

\begin{figure}[ht!]
\centering
\begin{emp}[classdiag](20,20)
Class.graph("Graph")("+verteci:Array[Vertex]", "+edgeCount:Int")();
Class.vertex("Vertex")("+id:Int", "+connections:Seq[Int]")();

leftToRight.top(30)(graph,vertex);

drawObjects(graph,vertex);

link(association)(graph.e -- vertex.w);
item(iAssoc)("*")(obj.se = vertex.w);

\end{emp}
\caption{Case classes}
\end{figure}

So the file is stored in an imutable memory structure. This is possible
because the connections are just the number that essently work as pointers
if you have the graph at hand.

\subsubsection{Solution generation}
Then I went on doing the generation of the solutions. I did this by creating
a random string of half the graph length, then copying this string and invert
it. This ensured there were as many ones as zeros. Then I combined these
two strings and randomly shuffled them:

\lstset{language=scala}
\begin{lstlisting}
	val random = Random
	def randomGene = random.nextInt(2).toString
	def randomGenes(geneLength:Int) = 1.to(geneLength).
		foldLeft(""){(b, x) => b + OffspringGenerator.randomGene}
	def randomBalancedGenes(geneLength:Int):String = {
		if(geneLength % 2 == 1){
			throw new Exception("Doesn't work with odd genelength")
		}
		val initial = randomGenes(geneLength/2)
		random.shuffle(
			inverseMirror(initial).toCharArray.toSeq
		).mkString
	}
	def flipBit(c:Char):Char = inverseResult(c,'1','0')
	def inverseMirror(input:String):String ={
		input + input.map(flipBit)
	}
\end{lstlisting}

\subsubsection{Valuation}
After that I did the evaluation function. Which was just folding the vertexes
into a float and then counting the string indexes that did match. I think
I did this wrong by mistake at first, but I realized it was advantageous to
count the connections that are the same instead of different. Because now
we want to maximize again as in the previous assignment, which made it
possible to leave the selection as is (so I don't have to modify that when
implementing GLS). To represent the internal valuation 
(which is trying to be maximized) as the minimizing
problem I simply do $graph.edgecount-valuation$.

\subsubsection{Local search}
Finally I designed the search class, which was first called LocalSearch but
was later generalized to Search:

\begin{figure}
\begin{emp}[classdiag](20,20)
Class.search("Search")(
	"-method:IMember=>IMember", 
	"-stopCondition:(IMember, IMember)=>Boolean",
	"-decideResult:(IMember, IMember)=>IMember"
	)(
	"+search(previous:IMember):IMember"
	);

drawObjects(search);
\end{emp}
\caption{Search}
\end{figure}

A very compact class indeed, we execute method on previous, check the
stopcondition if true we decide result else we call ourself (tail) recursively.
The stopcondition and decide result are by default equal and return the result
of the method. Why these functions are configurable will be discussed in the 
ILS chapter.
Then the method is either filled with vertexSwapFirstImprovement or
FiducciaMattheysesSearch.

@@

I got much better results with a system where the local search got several chances.
If it failed it would fall back, and then try again.
With chances around 30 I managed to get up and around 1900.
When I reset the chance counter on a successful find
I managed to get consistently 2200, one result was as high as 2310. Note:
I was maximizing here and still counting double. For the final run I put
the validation count up to 100. So before we decide something is at a local
optimum it first has to return the same result 100 times.

\subsubsection{Implementation}
At first I started to wrote a separate ILS class for search below the then
local search class. However I noticed the code I was producing looked 
suspiciously similar to what I already had in local search. So I decided
to try and combine the functionalities by making LocalSearch more configurable
and I succeeded. The end result was the Search class (I had to rename
the local search class because it was no longer just doing local search).

To explain how it works the first thing one should do is notice how the method
of search has the same prototype as the search method of the search class.
This means you can nest a search method of one search class inside another
search class. So we now have one inner search class doing the local search
and one outer search class doing iteration and perturbation of solutions.

Actually the perturbation is also nested inside a method like prototype and
I use function composition to let the perturbation be executed just before
the local search can be executed.

as stopcondition of the outer search we have an isWorse function inside a
resetretryOnBetter function. this function has also access to the retryOnResult
object which counts to the validation (and can be rest on a better result).
\subsection{GLS}
I implemented GLS with the evolution class. The only thing I had to do really
was write the crossover method, the rest was already functional.

@@

So I also started to think about my own topic to investigate, at first
I just wanted to do varying population sizes and varying selection methods.
But I thinks this takes a little to much time and it doesn't quite fit
on the tTest table so I decided to drop the varying pop-sizes.

\subsection{Counting double}
I had an issue in which my classmates would get much better results than me.
After discussing it for a while it turned out that I was counting double. However
This didn't make sense because I just used Scala combinators to count the
elements. This would mean that there should be a bug in the std
Scala library which seems highly unlikely. I knew I was counting double
because all my results were even numbers. After thinking about it for a while
I decided to investigate the graph file, it turns out that every connection
is made double, so the double counting was in fact correct.

\subsubsection{Graph pruning}
I decided to use a prune graph to speed up the algorithm. In the beginning I
create 2 graphs. One normal and one pruned, The pruning works by
comparing all connections against the id, if lower we ignore the connection.
To make sure we only do this on doubly linked graphs 2 graphs are created.
If the pruned graph has exactly half the score of a random solution we assume
its a doubly linked graph so we can prune it. I know its not a
guaranteed to work.
The pruning effectively halves the amount of connections that need to be
checked by each iteration (this happens in the evaluation). The
Fiduccia Mattheyses has even more gain because it has a lot of book keeping to
do with the connections.

\subsection{Fiduccia Mattheyses}
I was somewhat sceptical of my local search implementation. I had quite
different results from my classmates so to compare I decided to implement
the Fiduccia Mattheyses local search to. It was quite a lot of work but it
truly is significantly faster. Finding the paper was somewhat difficult
as the original was behind a paywall.  But someone put a copy on GitHub so
I just used that one.

The algorithm involves a lot of clever book keeping.
First of all the concept of gain is introduced, this indicates how much
gain there will be from moving a vertex to the other partition. Then the
vertexes are split based on partition and put into an array with as index
their gain (the paper calls this a bucket, as do I).
If there are more vertices at a particular index then they're
put into an linked list. To support this I introduced a container class
Cell that holds the vertex and also has an optional previous and next
element. (Again the paper uses the name cell to).

These cells are also put in another array based on their vertex ID (position
in the solution string). This
array is used for constant access speed when the \emph{connections of a vertex}
their gain needs to be updated after moving the initial vertex to the other
partition.
So you can just untie the element of the linked list (or if it has no
previous in my case, its in the front of the bucket and I remove it from
there and reinsert the next element and itself to the same bucket at the
proper gain index).

@@

As can be seen in the results section the Fiduccia Mattheyses algorithm is
much faster than first improvement vertex swap. The main reason for this is
that a single pass will climb very steeply in valuation. It also appears
to produce better results, I think this is because highest gain is considered
first. But more about this in the results.

\subsection{Statistics}
The error margins in the graph are the maximum range of the results so these
are round numbers when subtracted from the low or added to the high. The low
and high values are calculated with something called a confidence interval.
It aligns pretty well with the tTest results.

@@

\end{tikzpicture}
\\
\input{vertResults.textable}
\end{empfile}

\section{Conclusion}
Fiduccia Mattheyses combined with GLS finds the best solution at the cost of
quite some time, but the results are very close to the global optimum.

\end{document}
